A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-10-17 thesis/thesis.tex
fix Future Directions

\begin{document}
\title{Wait-free Queues with Polylogarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

@@

\pagebreak
\section{Future Directions}
We designed a tree to achieve agreement among processes in poly-log steps with respect to the total number of operations invoked and used it to implement a poly-log queue with respect to the size of the queue. We can use our solution for agreement to construct more poly-log wait-free linearizable objects. We call our solution \it{Block Tree}. In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.

\paragraph{Reducing Space Usage}
Our queue algorithm's usage of memory has places to improve. We assumed arrays in the nodes are unbounded.  If there is an $N$, such that the $N$ is greater than the  maximum possible operations appended to the block tree, then we can use Array
Segment~\cite{7073592} model to take $O(n+ \log N)$ space in each node, where $n$ is the total number of operations. To implement this we can create a static array of pointers called \nf{arr} of references to array segments. When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array with size $2^i$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access to $i$th element it looks up to \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]} which takes $O(1)$ steps. Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened. Furthermore at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]} which in average the time taken to allocate arrays in an execution is $O(1)$ per operation.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array Segments}
  \label{fig::doublingArray}
\end{figure}

\paragraph{Garbage Collection}
We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
We can keep track of the \nf{block}s in the \nf{root} that all of their operations are terminated, i.e, all of its enqueues have been dequeued and the response of all of its dequeues is computed. We call these blocks \it{Finished Blocks}. If we help the old operations to compute their response, then we can say if block $B$ is finshed then all blcoks before $B$ are also finished. Knowing the most recent finished block we can release the memory taken by these blocks. To throw the garbage in the blocks away we cannot use arrays. We need a datastructure that supports \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$. \nf{split(i)} is to remove all the indices less than \nf{i}. We can use Persistent Red Black Trees~\cite{DBLP:conf/podc/BashariW21} for this usage.

@@

\paragraph{Poly-logarithmic Wait-free Sequences}
Assume a data structure storing a sequence that has supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}. \nf{append(e)} adds \nf{e} to the sequence, \nf{get(i)} gets the \nf{i}th element in the sequence and \nf{index(e)} computes the position of appended element \nf{e} in the sequence. We can modify our queue to design such data structure.
If we do not distinct between \nf{Enqueue}s and \nf{Dequeue}s and replace \nf{DSearch} with Binary Search on all the \nf{root} in our algorithm, we achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.

\paragraph{Stacks}
There are two reasons block tree worked well to implement a queue. Firstly, to respond a \nf{Dequeue} we do not need to look at all the history of operations, if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} in $O(\log n)$ if we keep the number of enqueues and the size. Secondly, the operations needed to look to respond a the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue. It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.

>>>>

\begin{document}
\title{A Wait-free Queue with Polylogarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

@@

\pagebreak
\section{Future Directions}
We designed a tree to achieve agreement on an linearization of operations invoked by $p$ processes in an asynchronous model, which we will call \it{Block Tree}. We also implemented two queries to know information about the ordering agreed in the block tree. Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes. Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects. In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.

\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. We could instead use the vector of Feldman, Valera-Leon and  Damian Segment~\cite{7073592} model to use $O(n)$ space in each node, where $n$ is the total number of operations. Their vector creates an array called \nf{arr} of pointers to array segments. When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array with size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(p\log n)$ which if $n>>p$ we can ignore the $p$ factor. The vector implementation has also a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array Segments}
  \label{fig::doublingArray}
\end{figure}

\paragraph{Garbage Collection}
We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e, all of its enqueues have been dequeued and the responses of all of its dequeues have been computed. We call these blocks \it{finished blocks}. If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished. Knowing the most recent finished block in a node we can reclaim the memory taken by finished blocks. To throw the garbage in the blocks away we cannot use arrays (or vectors). We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$. Where \nf{split(i)} removes all the indices less than \nf{i}. We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/podc/BashariW21}.

@@

\paragraph{Poly-logarithmic Wait-free Sequences}
Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}. An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence. We can modify our queue to design such data structure. \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} with replacing the \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
 We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.

\paragraph{Stacks}
There are two reasons block tree worked well to implement a queue. Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size. Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue. It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.
